---
title: "Custom Lmest Functions"
author: "Diego Vitali"
output: 
  github_document:
    toc: true
date: "`r format(Sys.time(), '%d %B, %Y')`"
editor_options: 
  markdown: 
    wrap: 72
---
  
```{r setup, eval=TRUE,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
library(LMest)
library(dplyr)
library(ggplot2)
library(gridExtra)

### optional
library(kableExtra)
```





# Introduction

This document proposes a set of additions to the LMest package to support
convergence diagnostics, model output interpretation, and parallel estimation.
All functions are illustrated using the built-in `PSIDlong` dataset. Each
proposed function is self-contained: it takes a fitted `lmest` model object as
its primary input and returns results consistent with existing LMest conventions.

**Proposed additions:**

- `lmestSearch_plot()` — complements `lmestSearch()` by plotting the
  distribution of log-likelihood differences across random starts, providing a
  visual and quantitative assessment of whether `nrep` is large enough to
  reliably reach the global maximum.

- `get_observed_InitialStates_counts()` — returns the observed count of
  subjects assigned to each latent state at baseline via global (Viterbi)
  decoding, providing a proxy class-size check for separation and EPV
  diagnostics.

- `get_observed_transition_counts()` — returns the observed count of
  state-to-state transitions via global or local decoding, providing the
  denominator for EPV-based adequacy checks on transition probability estimates.

- `initial_prob_Diagnostics()` — summarises covariate coefficients for the
  initial state probabilities with standard errors, confidence intervals,
  significance flags, EPV, and separation warnings.

- `transition_Diagnostics_multilogit()` — summarises covariate coefficients
  for the transition probabilities from a specified starting state, with the
  same gatekeeping flags as `initial_prob_Diagnostics()`.

- `lmest_parallel()` — wraps `lmest()` in a `future_lapply()` call to
  distribute independent random-start replications across parallel workers,
  selecting the best model by log-likelihood and reporting a stability
  summary across workers.

---

```{r}

data("PSIDlong")
```


* Defining latent states indicators using "Fertility" and "employment".
* Defining predictors of initial states prob. a transition prob: In this case the predictors are the same for both.
```{r}

latent_state_vars<-c("Y1Fertility", #Y1
                     "Y2Employment" #Y2
                     )


initial_prob_vars<- c( "X1Race",       
                       "X2Age",        
                       "X4Education",
                       "X9Income"
                     )

trans_prob_vars <-   initial_prob_vars

```



# Plotting function for lmestSearch()

In this work I followed (as much as possible) Moore et al [(2025)][https://onlinelibrary.wiley.com/doi/full/10.1002/ijop.70021] and other papers mentioned in the pipeline. The basis lmestSearch() running all the possible K+ 1 models. Given the customary lmest example (PSIDlong), indicators are 2 binary variables, therefore we can set lmestsearch with k+1 = 5. 

## What lmestsearch does
The existing function lmestSearch() addresses model selection based on AIC and BIC and provides information to explore the multimodality of the likelihood function. Lmest provides two main criteria to select the number of latent states: AIC and BIC. For example, for the PSIDlong dataset we can estimate the basic LM model for increasing values of the latent states $k$. Therefore, the lmestSearch systematically fit models with various values of $k$ and compares their fit based on information criteria (AIC and BIC) as also suggested in Moore et al. (2025). The lmestSearhc function progresses as follow:

1. lmestSearch() runs the estimation for several values of $k$ (number of latent classes/states), each time recording log-likelihood ($lk_k$), AIC, and BIC.
2. For each value of $k$, lmestSearch() performs multiple random initializations to mitigate the risk of local maxima due to likelihood multimodality. This is explicitely warranted in Moore et al. (2025).
3. lmestSearch() outputs and compares AIC and BIC values across candidate models, so that we can select the solution that best balances fit and parsimony. BIC is generally preferred in longitudinal latent class modeling for its stronger penalty on model complexity, as per the recommendations in the literature.

Therefore here I aim to:

1. explore what is a reasonbly large number of tries
2. look at both lk and AIC/BIC to determine the optimal $k$


```{r results='hide',eval=TRUE,cache=T,warning=FALSE,message=FALSE,error=FALSE}
fmLatent<-lmestFormula(data = PSIDlong, response = latent_state_vars, 
                                LatentInitial = initial_prob_vars,
                                LatentTransition =trans_prob_vars)

Lmest_search_out <- lmestSearch(responsesFormula =  fmLatent$responsesFormula,
                   index = c("id","time"),
                   data = PSIDlong,
                   version ="categorical",  # type of response
                   k = 1:5,
                   modBasic = 1, 
                   nrep = 20,
                   maxit = 5000,
                  # fort = T,
                   seed = 12345)

summary(Lmest_search_out)

plot(Lmest_search_out$out.single[[2]], what ="CondProb")
plot(Lmest_search_out$out.single[[3]], what ="CondProb")
plot(Lmest_search_out$out.single[[4]], what ="CondProb")
plot(Lmest_search_out$out.single[[5]], what ="CondProb")

Lmest_search_out$out.single[[2]]$Psi
Lmest_search_out$out.single[[3]]$Psi
Lmest_search_out$out.single[[4]]$Psi


```


In lmest() the number of traces (fit attempts) per k is proportional to (k-1) and *nrep* so that number of random starts will be $ (k-1) * nrep$. In particular lmestSearch() proceed with 1 deterministic baseline initialisation followed by $ (k-1) * nrep$ random start, and then completes the exploration with a determininist refined solution that uses initial values taken from the best solution:


| Start Value | Purpose                                         | Typical Count per \(k\)  |
|-------------|------------------------------------------------|---------------------------|
| 0           | Deterministic baseline initialization          | 1 run                     |
| 1           | Random starts with random initial values       | \( nrep \times (k-1) \)   |
| 2           | Initial values are taken from best solution    | 1 run                     |

for example, for $k=3$, with $nrep = 20$ our lmestSearch output will have an lktrace array which carries the following 42 unit long lltrace sequence (Lmest_search_out$out.single[[3]]$lktrace):

* first ll: determininistic ll (start = 0). 
* 40 random starts: 20 * (3-1) = 40 random start (start = 1), 
* last ll: determininistic approach that (i think!) uses the paramenters of the best model as starting values(start = 2). 

the ll value with start =0 and that with start = 2 are respectively the first and the last value in each lktrace.

NOTE: **I took this from the lmest source code and by observing the output**

**If LMest ever changes the ordering of lktrace or adds entries, the slice [2:(n-1)] silently produces wrong results. this is a soft dependency on the internal structure. However, if LMest() in the future could expose lktrace with named elements this plotting function will become more robust  -- e.g. list(deterministic = .. ; random = ...; refined = ...).**

```{r }
# lktrace: all convergence ll
Lmest_search_out$out.single[[3]]$lktrace

# how many runs, in this case (start = 1) and k = [3] we have 40 random starts: 20 * (3-1) 
length(Lmest_search_out$out.single[[3]]$lktrace)

```

**BUT how many random starts is enough ?**

In the case of PSIDlong the reponse model is not very complex, but if we have a complex response model it maybe difficult to confidently chose a adequate number of random starts. The aim of this plot function is compliment lmestsearch and help me evaluate the number of random starts used: is *nrep* large enough to adequately explore the parameter space? is *nrep* large enough to reach a convergence ll parameter that is unlikely smaller than the global maximum - up to a negligeable tolerance level.

In order to help supporting this process I set up *lmestSearch_plot()*

## What lmestSearch_plot() adds to lmestsearch()

The aim of this function is to assess the stability and convergence of latent markov model fits by analysing the distribution of log-likelihood values obtained from multiple random starts.

The function takes as input a collection of log-likelihood traces obtained from different initialization runs for a given number of latent states (k). It focuses on the randomly initialized runs, excluding the initial deterministic run and the final refinement run. The aim here is to analyze the dispersion of the log-likelihood values found via the random starts.

By calculating the difference between each run’s log-likelihood and the highest log-likelihood observed across all runs, we capture how close each solution is to the best estimate. We then compute summary statistics like the median and the median absolute deviation (MAD) of these differences to understand the spread in the estimates.

To identify potentially unstable or suboptimal runs, the function uses a cutoff defined as the sum of the median difference plus a user-defined multiple of the MAD (defaulting to 3). Miller (1991) recommends using 2, 2.5, or 3 as the value k, depending on the purpose of outlier detection, while Leys et al. (2013) recommend a criterion of 2.5". Runs with log-likelihood differences exceeding this cutoff are flagged as possible outliers, indicating that these runs are likely converging to inferior local maxima. This is particularly helpful to visualise the distribution of the ll_diff scores that are closer to the best LL value.



```{r }
lmestSearch_plot <- function(all_lks, k = 4, k_multiplier = 3, plot_hist = TRUE) {
  best_lk <- all_lks[[k]]$lk
  lktrace_all <- all_lks[[k]]$lktrace
  
  # Assume that lktrace contains:
  #  . one start=0 fit as first lk,
  #  . nrep*k-1 start = 1 lk
  #  . one start=2 fit as last run
  # I exclude both start0 and start2 to focus on the random start distribution
  if(length(lktrace_all) > 2){
    random_lktrace <- lktrace_all[2:(length(lktrace_all) - 1)]
  } else {
    random_lktrace <- lktrace_all
  }
  
  # Store absolute best (probably from start=2, last run)
  # perhaps here we could compare with last run to flag whenever the absolute 
  # best lk is not coming from start=2
  abs_best_lk <- lktrace_all[length(lktrace_all)]
  
  max_lk <- max(random_lktrace)
  diff_array <- random_lktrace - max_lk
  abs_diff <- abs(diff_array)
  
  median_diff <- median(abs_diff)
  mad_diff <- mad(abs_diff)
  # Cutoff calculation
  # Define cutoff as median + 3 * MAD [https://pmc.ncbi.nlm.nih.gov/articles/PMC8801745/] 
  # "For example, Miller (1991) recommends using 2, 2.5, or 3 as the value k, depending on 
  # the purpose of outlier detection, while Leys et al. (2013) recommend a criterion of 2.5 as the value k"
  cutoff <- median_diff + k_multiplier * mad_diff
  
  filtered_diffs <- diff_array[abs_diff <= cutoff]
  dropped_diffs <- diff_array[abs_diff > cutoff]
  drop_ratio <- length(dropped_diffs) / length(abs_diff)
  
  df_all <- data.frame(diffs = diff_array, type = "Random starts")
  df_filtered <- data.frame(diffs = filtered_diffs, type = "Filtered")
  
  if(plot_hist){
    p1 <- ggplot(df_all, aes(x = diffs)) +
      geom_histogram(fill = "lightblue", color = "black", bins = 50) +
      geom_vline(xintercept = abs_best_lk - max_lk, color = "red", linetype = "dashed") +
      labs(title = paste0("Random starts log-lik diffs for k=", k),
           x = "Log-likelihood difference from max",
           y = "Count",
           caption = "Red dashed line: absolute best log-likelihood found (start=2 refinement run)") +
      theme_minimal()
    
    p2 <- ggplot(df_filtered, aes(x = diffs)) +
      geom_histogram(fill = "lightblue", color = "black", bins = 50) +
      geom_vline(xintercept = abs_best_lk - max_lk, color = "red", linetype = "dashed") +
      labs(title = paste0("Filtered log-lik diffs for k=", k, 
                          " (tol = median + ", k_multiplier, "*MAD)"),
           x = "Log-likelihood difference from max",
           y = "Count",
           caption = paste0("Red dashed line: absolute best log-likelihood found (start=2 refinement run). \n Drop ratio: ", round(drop_ratio, 3))) +
      theme_minimal()
    
    grid.arrange(p1, p2, ncol = 1)
  }
  
  return(list(
    median_diff = median_diff,
    mad_diff = mad_diff,
    cutoff = cutoff,
    drop_ratio = drop_ratio,
    filtered_diffs = filtered_diffs,
    dropped_diffs = dropped_diffs,
    abs_best_lk = abs_best_lk
  ))
}
```

In this example we examine the random starts for k = 2,3,4,5

```{r}

state2_llDiffs<-lmestSearch_plot(Lmest_search_out$out.single,k = 2, k_multiplier = 3, plot_hist = TRUE)
state3_llDiffs<-lmestSearch_plot(Lmest_search_out$out.single,k = 3, k_multiplier = 3, plot_hist = TRUE)
state4_llDiffs<-lmestSearch_plot(Lmest_search_out$out.single,k = 4, k_multiplier = 3, plot_hist = TRUE)
state5_llDiffs<-lmestSearch_plot(Lmest_search_out$out.single,k = 5, k_multiplier = 3, plot_hist = TRUE)

cat("Median absolute diff near max for state 2:", abs(state2_llDiffs$median_diff), "\n")
cat("Median absolute diff near max for state 3:", abs(state3_llDiffs$median_diff), "\n")
cat("Median absolute diff near max for state 4:", abs(state4_llDiffs$median_diff), "\n")
cat("Median absolute diff near max for state 5:", abs(state5_llDiffs$median_diff), "\n")



```

The plots complement the output of lmestSearch() and allow to evaluate the stability of the estimate, they expose multimodality of the log-likelihood difference providing valuable insight on both the stability of the convergence at each k state, and the complexity of the parameter space.



# LMest() output handlers, basic gatekeeping and diagnostics

I propose here some tools that can facilitate some aspects of working with lmest():

* get_observed_InitialStates_counts()
* get_observed_transition_counts()
* initial_prob_Diagnostics()
* transition_Diagnostics_multilogit()



```{r warning=FALSE}
mod2 <- lmest(responsesFormula = fmLatent$responsesFormula,
             latentFormula =  fmLatent$latentFormula,
             index = c("id","time"),
             data = PSIDlong, 
             k = 3, # diagnostics suggest 2 but I need 3 for a later example on parameter estimates
             paramLatent = "multilogit",
             start = 0, 
             seed = 1234,
             out_se=TRUE) 

plot(mod2, what="CondProb")


```

## get_observed_InitialStates_counts()

*it is important that the model solution is interpretable, which includes having multiple characteristics. First, from an absolute sense, the sample size for the smallest class needs to be large enough for the researcher to be comfortable interpreting it out to the population. Class size can become a challenge, particularly if there are realistically low frequency classes enumerated. For example, if examining classes related to types of drug use, the proportion of responders who regularly use heroin is likely small in both the population and the representative sample. *
[https://onlinelibrary.wiley.com/doi/full/10.1002/ijop.70021]

The *get_observed_InitialStates_counts()* uses global decoding to get the "cell sizes" for each initial latent state. I use global decoding as default here to get the "$Ug$ object. The first column of this object will be "time 1" state allocation for all subjects in the data set.

These counts are essential because (provided a decent model fit) these decoded class membership counts should be (very) close to the expected membership targetted by the multinomial logistic regression. Of course the decoded states does not capture the uncertainty that exists in the model around state membership but here I assume they provide a proxy base for diagnostics. *Please advise if this assumption is wrong.*

```{r,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
get_observed_InitialStates_counts <- function(model) {
  # Ul is a matrix of dimensions [n_subjects, n_timepoints]
  # We want the first time point (initial states)
  dec <- lmestDecoding(model)

  # Choose decoding type: "global" (Viterbi) or "local"
  # For global:
  initial_states <- dec$Ug[, 1]  # First column = time 1
  
  # Count observations in each state
  state_counts <- table(initial_states)
  
  # Convert to named vector with proper labels
  state_names <- paste("State", 1:model$k)
  N_obs_states <- setNames(as.numeric(state_counts), paste("State", names(state_counts)))
  
  # Ensure all states are represented even if count is 0
  all_states <- setNames(rep(0, model$k), state_names)
  all_states[names(N_obs_states)] <- N_obs_states
  
  return(all_states)
}
```

### example usage
```{r}
N_obs_states<-get_observed_InitialStates_counts(mod2)

N_obs_states %>% kable(col.names = c("","count"), caption = "Count of respondents by each allocated latent state at baseline")
```


## get_observed_transition_counts()

This is very similar to get_observed_initialStates_counts() but with the transition counts I put the option to chose whether using global decoding or local deconding. Exacly as we did for the initial state counts, the transition counts estimated via global decoding are essential in this function because they provide a proxy base for many diagnostics.


```{r}
get_observed_transition_counts <- function(model, option = "global") {
  # Get decoded sequences
  dec <- lmestDecoding(model)

  if (option == "global") {
    cat("Using global decoding: select option = local for local decoding ")
    latseq <- dec$Ug  # global decoding
  } else if (option == "local") {
    cat("Using local decoding: select option = global for global decoding ")
    latseq <- dec$Ul  # local decoding
  } else {
    stop("Invalid option. Use 'global' or 'local'.")
  }
  
  k <- model$k
  transition_counts <- matrix(0, nrow=k, ncol=k)
  
  for (i in 1:nrow(latseq)) {
    for (t in 1:(ncol(latseq)-1)) {
      from <- latseq[i, t]
      to   <- latseq[i, t+1]
      if (!is.na(from) && !is.na(to)) {
        transition_counts[from, to] <- transition_counts[from, to] + 1
      }
    }
  }
  
  rownames(transition_counts) <- paste0("From State ", 1:k)
  colnames(transition_counts) <- paste0("To State ",   1:k)
  
  return(transition_counts)
}

```

### example usage
```{r}
Nobs_trans<-get_observed_transition_counts(mod2, option = "global")
# Global decoding gives a robust and realistic count of transitions that can be helpful for diagnostic purposes
print(Nobs_trans)# %>% addmargins())
```

## initial_prob_Diagnostics()

* requires N_obs_states which can be obtain via get_observed_InitialStates_counts()

*initial_prob_Diagnostics()* summarizes and diagnoses the covariate effects on initial state probabilities in a latent Markov model fitted with LMest. The function assumes that LMest takes "state 1" as the relative reference state for each logistic regression. The function extracts, for each chosen covariate and each latent state the estimated regression coefficient, standard error, t-value, and confidence intervals. It also calculates "Events Per Variable" (EPV) for each combination, which gives the reader a sense of the adequacy of the sample size for each estimate.

To aid interpretation and flag potential model or data issues, the function adds some basic gatekeeping:

* A significance code (stars), based on the t-value, corresponding to standard thresholds.
* Diagnostics for possible separation or estimation instability, including flags for extreme coefficients, high standard errors, low EPV, and states with few observed cases.
* Human-readable comments to further contextualize the estimate quality and highlight warnings (e.g., for overfitting, separation, or low precision).

The output is a tibble with key statistics and quality checks for each covariate-state combination. I intended this for robust reporting in my analysis and to facilitate interpretation of initial probability regression results of my latent Markov models.
```{r}
initial_prob_Diagnostics <- function(model, var_names = NULL, N_obs_states, conf_level = 0.95, skip_intercept = TRUE) {
  # Get number of states
  k <- model$k
  
  # Get covariate names from Be dimensions (these are the initial prob regressions)
  all_covars <- dimnames(model$Be)[[1]]
  
  # If var_names is NULL, use all variables (optionally excluding intercept)
  if (is.null(var_names)) {
    if (skip_intercept && "(Intercept)" %in% all_covars) {
      var_names <- all_covars[all_covars != "(Intercept)"]
    } else {
      var_names <- all_covars
    }
    covariate_indices <- match(var_names, all_covars)
  } else {
    # Get indices for the variables of interest
    covariate_indices <- match(var_names, all_covars)
    
    # Check for invalid variable names
    if (any(is.na(covariate_indices))) {
      invalid_vars <- var_names[is.na(covariate_indices)]
      message("Variables requested:", paste(var_names, collapse = ", "))
      message("Variables available in model:", paste(all_covars, collapse = ", "))
      stop(paste("Variable(s) not found in model:", paste(invalid_vars, collapse = ", ")))
    }
  }
  
  # Get coefficient and SE matrices
  coeff_mat <- model$Be
  se_mat <- model$seBe
  
  # Critical value for confidence intervals (hardcoded as two tailed)
  z_crit <- qnorm((1 + conf_level) / 2)
  
  # States 2, 3, .., k (assuming state 1 is reference)
  states <- 2:k
  state_labels <- paste("State", states)
  
  # Get N_obs for each state from provided vector
  N_obs_vec <- N_obs_states[state_labels]
  
  # Build dataframe for each state
  df_list <- lapply(seq_along(states), function(i) {
    state <- states[i]
    ref_state<- 1
    col_idx <- i  # Column index in Be/seBe matrices (columns are "2", "3", "4")
    
    coef_val <- coeff_mat[covariate_indices, col_idx]
    se_val <- se_mat[covariate_indices, col_idx]
    
    data.frame(
      varname = var_names,
      State = paste("State", state),
      Ref_state = paste("State", ref_state),
      N_obs_states = N_obs_vec[[i]],
      Coefficient = round(coef_val, 3),
      StdError = round(se_val, 3),
      t_value = round(coef_val / se_val, 2),
      CI_lower = round(coef_val - z_crit * se_val, 3),
      CI_upper = round(coef_val + z_crit * se_val, 3),
      stringsAsFactors = FALSE
    )
  })
  
  # Combine all dataframes
  df <- do.call(rbind, df_list)
  ####################################################################
  ############  GATEKEEPING AND DIAGNOSTICS
  ####################################################################
  # Add EPV (Events per parameter)
  # For initial probabilities, EPV = number in state / number of parameters
  p_params <- dim(model$Be)[1]  # Number of parameters per logit
  df$EPV <- round(df$N_obs_states / p_params, 2)
  
  # Add significance and separation flags
  df <- df %>% 
    mutate(
      Signif = case_when(
        ## based on conventional t_value assumptions (two-sided)
        ## in non-normal settings these thresholds might not strictly hold
        abs(t_value) > 2.58 ~ "***", # ~ p < .01
        abs(t_value) > 1.96 ~ "**",  # ~ p < .05
        abs(t_value) > 1.64 ~ "*",   # ~ p < .10
        TRUE ~ ""
      ),
      ## these are some separation sanity check
      Sep_Flag = case_when(
        abs(Coefficient) > 10 ~ "!!!",
        abs(Coefficient) > 5 & StdError > 2 ~ "!!",
        abs(Coefficient) > 5 ~ "!",
        EPV < 10 ~ "EPV < 10 !!",
        ## anything below EPV should undergo shrinkage should be applied
        # https://www.sciencedirect.com/science/article/pii/S0895435616300117
        EPV < 20 ~ "EPV < 20 !",
        StdError > 5 ~ "Extreme SE!",
        TRUE ~ "OK"
      ),
      Comment = case_when(
        is.infinite(Coefficient) ~ "WARNING: Inf. coef - compl. separation",
        Sep_Flag == "!!!" ~ "WARNING: Separation likely",
        Sep_Flag == "!!" ~ "WARNING: Quasi-separation",
        Sep_Flag == "!" ~ "WARNING: Very large coef",
        StdError > 5 ~ "WARNING: Extreme SE!",
        EPV < 10 ~ "WARNING: EPV < 10 overfitting",
        N_obs_states < 20 ~ "WARNING: < 20 obs in state",
        N_obs_states < 30 ~ "WARNING: < 30 obs in state",
        EPV < 20 ~ "Low EPV: coef requires shrinkage", # https://www.sciencedirect.com/science/article/pii/S0895435616300117
        StdError > 1 ~ "CAUTION: SE > 1",
        abs(t_value) >= 2 & StdError <= 1 ~ "Reliable",
        abs(t_value) >= 1.6 & StdError <= 1 ~ "Marginal",
        abs(t_value) < 1.6 & StdError <= 1 ~ "n.s",
        TRUE ~ "Check_manually"
      )
    )
  
  # Reorder columns for better readability
  df <- df %>% 
    select(varname, State, Ref_state, N_obs_states, Coefficient, StdError, t_value, 
           CI_lower, CI_upper, EPV, Sep_Flag, Signif, Comment) %>% 
    as_tibble()
  
  return(df)
}

```


### example usage

```{r}

## we need an array of the variable names that were used as predictors
# to be on the safe side we can get this from the initial probability model itself
Be_varnames<-(mod2$Be %>% row.names())[-1]
Be_varnames

# # Or for specific covariates by name
initial_States_summary<- initial_prob_Diagnostics(
   model = mod2,
   # if you omit varnames it will extract coef for all variables
   #var_names  = c("X1Race","X2Age"), 
   N_obs_states = N_obs_states,
   # "skip intercetp omits the intercept i.e. the log-odds of being in 
   # state 2,3,4 versus 1 when all predictors are zero.
   skip_intercept = TRUE, #  I did not need this but may be of interest.
   conf_level = 0.95
   
 )
initial_States_summary %>% 
  arrange(Sep_Flag,varname,State) 
 
```

## transition_Diagnostics_multilogit()

Requires N_obs_trans, the observed counts of transitions between latent states, a proxy of these counts can be obtained via global decoding using get_observed_transition_counts().

*transition_Diagnostics_multilogit()* summarizes and help diagnoses the covariate effects on transition probabilities from a specified start latent state in a latent Markov model fitted with LMest using multinomial logistic regression.

From the source code of lmcovlatent() LMest uses the "staying in the same state" transition as the baseline category in the multinomial logistic regression for transition. Consequently the model extracts, for each chosen covariate and destination state (excluding the reference state), the corresponding regression coefficient, standard error, t-value, and confidence intervals at a given confidence level (default 95%).


Diagnostics included:

* "Events Per Variable" (EPV) based on observed transition counts (this is the proxy count obtained via global decoding) and the number of covariates

* Separation Flags based on:
  - absolute logit coefficients larger than 5
  - extremely large SE (> 5)
  - convergence failure or warnings
  - low transition counts (sample size needs to be evaluated across all transitions)


EPV provides good insight into the adequacy of the sample size relative to the model complexity for each estimate. In fact, low transition counts can cause overfit and therefore biasing the predictors' estimate (e.g. causing inflation). Simulation studies of prediction models with binary outcomes suggested that as a rule of thumb that we should observe at least 10 EPV, sometimes 20+, to avoid overfitting and unstable coefficient estimates. In fact it is suggested that for EPV smaller than 20 "shrinkage" of regression coefficients should be applied to reduce the chance of overfitting. The diagnostics Flags reflect these suggestions [https://onlinelibrary.wiley.com/doi/full/10.1002/sim.8063]

I also added some general guidance of the precision of estimates (based on SE). Among the transitions that are unlikely affected by separation or quasi-separation, we can look at t-value (|Coefficient|/SE) to assess the general significance of the coefficient.


* |Coefficient|/SE < 1  : generally unreliable 
* 1 < |Coefficient|/SE < 2  : could be marginally reliable
* |Coefficient|/SE > 2  : generally reliable

Diagnostics flags are applied hierarchically via "case_when". The order of this hierarchy is debatable: I chose one. 

```{r}

transition_Diagnostics_multilogit<- function(model, start_state, covariates_of_interest, N_obs_trans, conf_level = 0.95) {
  k <- model$k
  all_states <- 1:k
  # here we get the order of destination states expected in the output of lmest at $Ga[,,start_state]
  dest_states <- setdiff(all_states, start_state) 
  # Map correctly - find Ga column index for each destination state
  col_idx <- match(dest_states, all_states[-start_state])
  
  all_covars <- dimnames(model$Ga)[[1]]
  
  if (is.null(all_covars)) {
    var_names <- covariates_of_interest
  } else {
    covariate_indices <- if (is.numeric(covariates_of_interest)) {
      covariates_of_interest
    } else {
      match(covariates_of_interest, all_covars)
    }
    var_names <- all_covars[covariate_indices]
  }

  coeff_mat <- model$Ga[, , start_state]
  se_mat <- model$seGa[, , start_state]

  # Critical value for confidence intervals (hardcoded as two tailed)
  z_crit <- qnorm((1 + conf_level) / 2)
  from_label <- paste("From State", start_state)
  dest_labels <- paste("To State", dest_states)
  
  N_obs_vec <- N_obs_trans[from_label, dest_labels]
  
  df_list <- lapply(seq_along(dest_states), function(i) {
    dest <- dest_states[i]
    coef_val <- coeff_mat[covariates_of_interest, col_idx[i]]
    se_val   <- se_mat[covariates_of_interest, col_idx[i]]
    varname  <- var_names
    
    data.frame(
      varname = varname,
      Transition = paste(start_state, "->", dest),
      N_obs_trans = N_obs_vec[[i]],
      Coefficient = round(coef_val, 3),
      StdError = round(se_val, 3),
      t_value = round(coef_val / se_val, 2),
      CI_lower = round(coef_val - z_crit * se_val, 3),
      CI_upper = round(coef_val + z_crit * se_val, 3),
      stringsAsFactors = FALSE
    )
  })
  
  df <- do.call(rbind, df_list)
  
  # Add EPV (Events per parameter) https://onlinelibrary.wiley.com/doi/full/10.1002/sim.8063
  p_per_logit <- dim(model$Ga)[1]  # same covars for each logit
  df$EPV <- round(df$N_obs_trans / p_per_logit,2)
  
  df <- df %>% 
    mutate(
      Signif = case_when(
        abs(t_value) > 2.58 ~ "***",
        abs(t_value) > 1.96 ~ "**",
        abs(t_value) > 1.64 ~ "*",
        TRUE ~ ""
      ),
      Sep_Flag = case_when(
        abs(Coefficient) > 8 ~ "!!!",
        abs(Coefficient) > 5 & StdError > 2 ~ "!!",
        abs(Coefficient) > 5 ~ "!",
        EPV < 10 ~ "EPV < 10 !!", # very low Event Per Variable: poor power, estimates will be unreliable: overfitting
        EPV < 20 ~ "EPV < 20 !", # Low Event Per Variable: still low power, estimates need shrinking to avoid overfitting
        StdError > 5 ~ "Extreme SE!",  # Unstable estimates
        TRUE ~ "OK"
      ),
      Comment = case_when(
        is.infinite(Coefficient) ~ "WARNING: Inf. coef - compl. separation",
        Sep_Flag == "!!!" ~ "WARNING: Separation",
        Sep_Flag == "!!"  ~ "WARNING: Quasi-separation",
        Sep_Flag == "!"   ~ "WARNING: Extreme coef",
        StdError > 5      ~ "WARNING: Extreme SE!",
        EPV < 10          ~ "WARNING: EPV < 10 overfitting",
        N_obs_trans < 20  ~ "WARNING: < 20 obs",
        N_obs_trans < 30  ~ "WARNING: < 30 obs",
        EPV < 20          ~ "Low EPV: coef requires shrinkage",
        StdError > 1      ~ "CAUTION: SE > 1",
        abs(t_value) >= 2 & StdError <= 1 ~ "Reliable",
        abs(t_value) >= 1.6 &  StdError <= 1 ~ "Marginal",
        abs(t_value) < 1.6 & StdError <= 1 ~ "n.s",
        TRUE ~ "Check_manually"
      )
    )
  # Reorder columns for better readability
  df <- df %>% 
    select(varname,Transition, N_obs_trans, Coefficient, StdError, t_value, 
           CI_lower, CI_upper, EPV,Sep_Flag, Signif, Comment) %>% as_tibble()
  
  return(df)
}

```

### Example usage

```{r}

Nobs_trans<-get_observed_transition_counts(mod2, option = "global")

Allvarnames<-(mod2$Ga[,,1] %>% row.names())[-1]

From_state1<-transition_Diagnostics_multilogit(mod2, 1, Allvarnames,Nobs_trans)
From_state2<-transition_Diagnostics_multilogit(mod2, 2, Allvarnames,Nobs_trans)
From_state3<-transition_Diagnostics_multilogit(mod2, 3, Allvarnames,Nobs_trans) 
summary_table <- bind_rows(From_state1, From_state2, From_state3)

summary_table %>% filter(EPV >= 20) %>% kable
summary_table %>% filter(EPV < 20) %>% kable

```

### Explaining how summary_table() parses LMest ouput arrays.

The following is an explanation on how this function pulls the summary tables. It is based on the output arrays of lmest, and its internals.

transition_Diagnostics_multilogit() produces a table that fetches the coefficients and standard error respecting the logic of which these are presented in $Ga[,,\bar{u}]$.

When I first started using LMest I was not sure about the meaning of the column numbering of $Ga[,,\bar{u}]$. The following paragraphs trace how I have investigated this to make sure my summary and diagnostic function was not misplacing coefficients and SE.

In the output of LMEST, object "Ga" is a three dimensional object
where $Ga[,,\bar{u}]$, $\bar{u},u = 1, \ldots, k$ shows the parameters for the transitions
starting from "state \bar{u}". 

E.g. $Ga[,,2]$ we see the parameter estimates for the transitions with starting 
point "state 2". The table shows intercepts and coeficients to calculate the log-odds of
transitioning to another state in comparison to staying in the same
state ($\pi_{\bar{u}|\bar{u}x}^{(t)}$). 

**NOTE** The column number in the table provided by LMest do not identify the destintation state "u".

In the example of  $Ga[,,2]$, the parametrisation uses as reference the
transition 2-\>2, and then:

- the 1st column (numbered as "1" in the output) shows coefficients for transition 2-\>1
- *(transition 2-\>2 is reference and omitted)*
- the 2nd column (numbered as "2") shows coefficients for transition 2-\>3

The above is not very intuitive but it is consistet. We can verify this by looking at the two specific dimensions of the $ga$ object: $ga[,m,]$ and $ga[,,\bar{u}]$.

$ga[,,\bar{u}]$ contains the values for the transition state 2-\>3, these are in the second column (column "2" in the LMest output)  and column 2 of the matrix $ga[ , ,2]$:
```{r}
mod2$Ga[,,2] %>% kable() # 2->3 coeffients can be found in 2 column
```


The $ga[,m,]$ matrix contains the parameter estimates as rows, and the starting state as columns $\bar{u}$. 

Therefore  $Ga[,2,2]$ is the coefficient vector for the *second* contrast (e.g., transition 
2->3 from starting *state 2*. Looking at this matrix helped me confirming that the column numbering of LMest does not return the "destination" state but just the "order" of estimation.

Why Ga[,2,] will contain the coefficients for transition 2->3? From the output and the code I assume $m$ is simply the order of the contrasts. There are $k-1$ of these $ga[,m,]$ matrices, so that (e.g.) $ga[,2,]$ shows the intercepts and coeficients for all the transitions that appear in the "second" column of $Ga[,,2]$. 

Therefore, as the column number of $ga[,m,]$ identify the starting state, in $Ga[,2,][2]$ we can find the coeficients for the transition 2->3 that we see in the LMest output $Ga[,,2][,2]$

```{r}
mod2$Ga[,2,2] %>% kable() # 2->3 coeffients can be found in 2 column
```

---

# Parallel multi-start fitting: `lmest_parallel()`

## Theoretical Motivation

A unique aspect of all finite mixture models is that although there is a global,
best-fitting solution at the global maximum log likelihood (LL), there are also
multiple less well-fitting solutions at different local maxima LL. Part of the
exploratory nature of finite mixture models is running each K+1 class model with
multiple random starts so that each model run includes multiple model solutions
with their LL reported in the output (Moore et al., 2025,
<https://onlinelibrary.wiley.com/doi/full/10.1002/ijop.70021>).

Well aware of this, `lmest()` already supports multiple random starts via `ntry` 
and `start = 1`. However, because the internal EM loop is single-threaded, all 
starts run sequentially. For complex models (many states, many covariates, large n),
achieving adequate coverage of the likelihood surface — i.e. replicating the
top LL multiple times from independent starts — requires a large total number
of starts, which can be prohibitively slow.

`lmest_parallel()` wraps `lmest()` in a `future_lapply()` call to distribute
`n_reps` independent `lmest()` calls across `n_workers` parallel processes.
Each worker runs a full `lmest()` call with `ntry` random initialisation and
returns its best model. The overall best model (highest LL across all workers)
is selected as the final result. A convergence summary and optional LL histogram
allow assessment of stability: if many workers independently recover the same LL,
it provides evidence of proximity to the global maximum.

**Dependencies:** `future` and `future.apply` (CRAN).

## Parameters

| Argument            | Default | Description |
|---------------------|---------|-------------|
| `responsesFormula`  | —       | Passed directly to `lmest()`. Defines the manifest response variables. |
| `latentFormula`     | `NULL`  | Passed to `lmest()`. Covariates on initial probabilities and transitions. |
| `index`             | —       | Character vector of length 2: subject ID and time variable names. |
| `data`              | —       | Long-format data frame. |
| `k`                 | —       | Number of latent states. |
| `n_workers`         | `detectCores() - 1` | Number of parallel processes to launch. Should not exceed available CPU cores. |
| `n_reps`            | `20`    | Total number of independent `lmest()` calls (replications). Each gets a unique seed. If `n_reps > n_workers`, `future` schedules them in batches — e.g. `n_reps = 12`, `n_workers = 4` → 3 batches of 4 parallel runs. |
| `ntry`              | `5`     | Number of random starts *per class* within each `lmest()` call. `lmest()` initialises each of the `k` classes independently, so total random starts = `n_reps × ntry × k`. |
| `base_seed`         | `1234`  | Starting seed. Worker `i` receives `base_seed + i`, ensuring full reproducibility. |
| `modBasic`          | `1`     | Passed to `lmest()`. `1` = time-homogeneous transitions. |
| `maxit`             | `10000` | Maximum EM iterations per start. |
| `tol`               | `1e-8`  | EM convergence tolerance. |
| `fort`              | `TRUE`  | Use Fortran routines (recommended for speed). |
| `out_se`            | `FALSE` | Compute standard errors. Leave `FALSE` during convergence search; refit the best model with `out_se = TRUE` separately.|
| `stability_tol`     | `5`     | LL units within which a replication is considered to have recovered the global maximum. Used for stability reporting only — does not affect model selection. |
| `plot_hist`         | `TRUE`  | If `TRUE`, builds a histogram of LL differences across all random starts (via `$lktrace`). The plot is returned in `$plot`; it is not printed by the function. |

### Batching

`future_lapply()` handles scheduling automatically. With `n_reps = 12` and `n_workers = 4`:

- 3 sequential batches, each running 4 workers in parallel
- total wall time ≈ time of one `lmest()` call × 3 batches (plus overhead)
- total random starts = 12 × `ntry`

With `n_reps ≤ n_workers` all replications run in a single batch.

### Return value

A named list:

| Element | Description |
|---------------------|---|
| `$best_model`       | The `lmest` object with the highest LL across all replications. Identical interface to a single `lmest()` call. |
| `$lks`              | Numeric vector of best LL per replication (length `n_reps`). |
| `$all_lls`          | Numeric vector of per-start LLs extracted from `$lktrace` (requires `lmest()` to expose `$lktrace` — currently empty). |
| `$all_diffs`        | `best_lk_all - all_lls`: LL gap of every random start from the global best (empty until `$lktrace` is available). |
| `$n_at_max`         | Number of workers whose best LL is within `stability_tol` of the global best. One value per worker — does not reflect all starts within each worker. |
| `$n_at_max_all`     | Number of individual random starts within `stability_tol` of the global best LL (requires `$lktrace`). |
| `$stability_pct`    | `n_at_max / n_reps × 100`: replication-level stability percentage. |
| `$duration_mins`    | Total wall time in minutes. |
| `$plot`             | A `ggplot2` object (LL difference histogram), or `NULL` if `plot_hist = FALSE`. Print with `print(out$plot)`. |

## Sketch

```{r eval=FALSE}
lmest_parallel <- function(
    responsesFormula,
    latentFormula   = NULL,
    index,
    data,
    k,
    n_workers     = parallel::detectCores() - 1,
    n_reps        = 20,
    ntry          = 5,
    base_seed     = 1234,
    modBasic      = 1,
    maxit         = 10000,
    tol           = 1e-8,
    fort          = TRUE,
    out_se        = FALSE, # no need to estimate these on all runs.
    stability_tol = 5,
    plot_hist     = TRUE
) {
  requireNamespace("future",       quietly = TRUE)
  requireNamespace("future.apply", quietly = TRUE)

  future::plan(future::multisession, workers = n_workers)
  on.exit(future::plan(future::sequential), add = TRUE)

  cat(sprintf(
    "lmest_parallel: k=%d, n_RandTry=%d via %d ntry x %d k in %d reps. Job allocated to %d workers (CPU cores)\n",
    k, n_reps * ntry * k, ntry, k, n_reps, n_workers
  ))

  t0 <- Sys.time()

  all_results <- future.apply::future_lapply(seq_len(n_reps), function(rep_id) {

    args <- list(
      responsesFormula = responsesFormula,
      index            = index,
      data             = data,
      k                = k,
      modBasic         = modBasic,
      start            = 1L,        # random starts
      ntry             = ntry,
      seed             = base_seed + rep_id,
      maxit            = maxit,
      tol              = tol,
      fort             = fort,
      out_se           = out_se,
      output           = FALSE      # suppress per-iteration output
    )
    if (!is.null(latentFormula)) args$latentFormula <- latentFormula

    capture.output({
      model <- tryCatch(
        do.call(LMest::lmest, args),
        error = function(e) NULL
      )
    })

    list(
      model  = model,
      lk     = if (is.null(model)) NA_real_ else model$lk,
      rep_id = rep_id,
      seed   = base_seed + rep_id
    )

  }, future.seed = TRUE)

  elapsed <- as.numeric(difftime(Sys.time(), t0, units = "mins"))

  # --- Select best model ---
  lks      <- sapply(all_results, `[[`, "lk")
  best_idx <- which.max(lks)
  best_lk  <- lks[best_idx]

  # --- Per-worker convergence summary ---
  lk_diffs  <- best_lk - lks
  n_at_max  <- sum(lk_diffs <= stability_tol, na.rm = TRUE)
  n_valid   <- sum(!is.na(lks))
  stab_pct  <- 100 * n_at_max / n_valid

  cat(sprintf("\nCompleted in %.1f minutes\n", elapsed))
  cat(sprintf("Best LL:              %.4f  (rep %d, seed %d)\n",
              best_lk,
              all_results[[best_idx]]$rep_id,
              all_results[[best_idx]]$seed))
  cat(sprintf("Stability (replications): %d / %d within %.0f LL units  (%.1f%%)\n",
              n_at_max, n_valid, stability_tol, stab_pct))
  if (n_valid < n_reps)
    warning(sprintf("%d worker(s) returned an error and were excluded.",
                    n_reps - n_valid))

  # --- Aggregate per-start LLs from lktrace across all workers ---
  # lktrace is a positional vector: [deterministic, random_1...random_n, refined]
  # Trim first and last entries to keep only random starts,
  # matching the same assumption made in lmestSearch_plot().
  # NOTE: if LMest exposed lktrace as a named list this trim would not be needed.
  all_lls <- unlist(lapply(all_results, function(x) {
    tr <- x$model$lktrace
    if (is.null(tr) || length(tr) <= 2) return(NULL)
    tr[2:(length(tr) - 1)]
  }))

  best_lk_all  <- max(all_lls, na.rm = TRUE)
  all_diffs    <- best_lk_all - all_lls
  n_at_max_all <- sum(all_diffs <= stability_tol, na.rm = TRUE)

  cat(sprintf("Stability (all starts):  %d / %d within %.0f LL units  (%.1f%%)\n",
              n_at_max_all, length(all_lls), stability_tol,
              100 * n_at_max_all / length(all_lls)))
  cat(sprintf("Total random starts captured: %d\n", length(all_lls)))

  # --- Optional histogram over all random starts ---
  p_hist <- NULL
  if (plot_hist && requireNamespace("ggplot2", quietly = TRUE)) {
    df_plot <- data.frame(lk_diff = all_diffs)
    p_hist <- ggplot2::ggplot(df_plot, ggplot2::aes(x = lk_diff)) +
      ggplot2::geom_histogram(fill = "lightblue", colour = "black", bins = 30) +
      ggplot2::geom_vline(xintercept = 0, linetype = "solid",
                          colour = "darkgreen", linewidth = 0.8)

    # Only draw tolerance line when it falls within the data range — otherwise
    # ggplot scales the x-axis out to stability_tol, compressing all bars left.
    if (stability_tol < max(all_diffs, na.rm = TRUE))
      p_hist <- p_hist +
        ggplot2::geom_vline(xintercept = stability_tol, linetype = "dashed",
                            colour = "red")

    p_hist <- p_hist +
      ggplot2::labs(
        title   = sprintf("LL differences from global best  (k = %d, %d workers, %d starts)",
                          k, n_reps, length(all_lls)),
        x       = "LL difference from global best (all random starts)",
        y       = "Count",
        caption = sprintf(
          paste0("Green solid line: global best LL = %.4f.\n",
                 "Stability tolerance: %.0f LL units",
                 "%s\n",
                 "Random starts only — deterministic and refinement starts",
                 " excluded via lktrace[2:(n-1)]."),
          best_lk_all, stability_tol,
          if (stability_tol >= max(all_diffs, na.rm = TRUE))
            " (all starts within tolerance — line omitted from plot)."
          else
            " (red dashed line).")
      ) +
      ggplot2::theme_minimal()
  }

  list(
    best_model    = all_results[[best_idx]]$model,
    lks           = lks,
    all_lls       = all_lls,
    all_diffs     = all_diffs,
    n_at_max      = n_at_max,
    n_at_max_all  = n_at_max_all,
    stability_pct = stab_pct,
    duration_mins = elapsed,
    plot          = p_hist
  )
}
```

## Notes for package integration

- **`output = FALSE`** suppresses the per-iteration EM print from each worker,
  keeping the console readable. You can re-enable this to TRUE for a more verbose mode
  e.g for single-rep debugging.

- **What this sketch omits relative to a fuller implementation:** in the
  analysis reported in tmy paper, I extended this sketch using `capture.output()` 
  around each `lmest()` call (run with `output = TRUE`) to parse three quantities 
  from the printed EM trace: 
    1. the final LL of each individual random start, enabling a per-start LL
  histogram analogous to `lmestSearch_plot()`; 
    2. the number of EM iterations each start required before convergence (`final_step`) -- a start converging
  in 8 iterations versus 9800 is diagnostically informative; 
    3. wall time per start. 
  Using capture.output was fine for running my pipeline once but obviously it is 
  fragile because it relies on the exact string format of LMest's internal `cat()` 
  calls and it would break silently if that format changed. 
  
  The iteration count and timing (points 2–3) have no equivalent field in the 
  current model object and would require LMest to expose them explicitly -- for 
  example as `$iter_per_start` and `$time_per_start`. Naming the entries of 
  `$lktrace` (e.g. `list(deterministic = lk0, random = c(...), refined = lkN)`) 
  rather than  returning a positional vector would also remove the structural 
  assumption that both `lmestSearch_plot()` as I suggested above in the search plot section. 

- **Memory:** each worker returns its full `lmest` model object, so `all_results`
  holds `n_reps` models in memory simultaneously. For large datasets this can be
  substantial (e.g. ~150 MB per model × 24 workers ≈ 3.5 GB). A leaner
  implementation would have workers return only `lk` and `lktrace` and recover
  the best model with a single refit using the winning seed:
  `lmest(..., seed = best_seed, ntry = ntry, start = 1)`.

- **Seed reproducibility:** each worker receives `seed = base_seed + rep_id`,
  making runs fully reproducible given the same `base_seed` and `n_reps`.

- **Error handling:** workers that fail (e.g. non-convergence, degenerate
  starting values) return `NA` for the LL and are excluded from the stability
  count with a warning.

- **`on.exit(plan(sequential))`** ensures the parallel backend is torn down even
  if the function exits with an error, avoiding leaked worker processes.

## Example usage

```{r eval=FALSE}
library(LMest)
library(future)
library(future.apply)

data("PSIDlong")

fmLatent <- lmestFormula(
  data             = PSIDlong,
  response         = c("Y1Fertility", "Y2Employment"),
  LatentInitial    = c("X1Race", "X2Age", "X4Education", "X9Income"),
  LatentTransition = c("X1Race", "X2Age", "X4Education", "X9Income")
)

# 12 replications on 4 cores -> 3 batches of 4 parallel runs
# total random starts = 12 reps x 5 ntry x 3 k = 180
out <- lmest_parallel(
  responsesFormula = fmLatent$responsesFormula,
  latentFormula    = fmLatent$latentFormula,
  index            = c("id", "time"),
  data             = PSIDlong,
  k                = 3,
  n_workers        = 4,
  n_reps           = 12,
  ntry             = 5,
  base_seed        = 12345,
  stability_tol    = 5,
  plot_hist        = TRUE
)

# Actual console output (PSID data, k=3, n_reps=4, ntry=3, n_workers=4):
# lmest_parallel: k=3, n_RandTry=36 via 3 ntry x 3 k in 4 reps. Job allocated to 4 workers (CPU cores)
# Completed in 13.5 minutes (Intel i7-10700, 2.9GHz)
# Best LL:                  -6575.8727  (rep 4, seed 12349)
# Stability (replications): 4 / 4 within 1 LL units  (100.0%)  <- best LL per worker only
# Stability (all starts):   0 / 0 within 1 LL units  (NaN%)  <- lmest() does not
# Total random starts captured: 0                             <- expose $lktrace;
#                                                                these lines require
#                                                                lmest() to store
#                                                                per-start LLs (as
#                                                                lmestSearch() does)

# Best model: drop-in replacement for a single lmest() call
out$best_model
plot(out$best_model, what = "CondProb")

# LL histogram — printed by caller, not by the function
print(out$plot)
```

